The hydra and the rooster

### 2020-07-21 latex/chapter-hydras.tex
corrections in the pdf documentation

Technically, a \emph{hydra} is just a finite ordered tree, each node of which 
has any number of sons. Note that, contrary to the computer science tradition, we will show the hydras 
with the heads up and the foot (i.e., the root of the tree) down.
Fig.~\ref{fig:Hy} represents such  a hydra, which will be referred to as \texttt{Hy} in our examples (please look at the module 
\url{../src/html/hydras.Hydra.Hydra_Examples.html}).

\begin{figure}[h]
\begin{tikzpicture}[very thick, scale=0.6]

@@

\begin{center}
\begin{tabular}{ll}
Hydras & Finite rooted trees\\
\hline
foot & root\\
head & leaf\\
node & node\\
segment  & (directed) edge \\
sub-hydra & subtree\\
daughter & immediate subtree\\
\end{tabular}
\captionof{table}{\label{tab:hyd2tree}
Translation from hydras to trees}
\end{center}

@@

\label{sec:orgheadline44}
A \emph{hydra battle} is a fight between Hercules and the Hydra. 
More formally, a  battle is a sequence of \emph{rounds}.
At each round:
\begin{itemize}
\item If the hydra is composed of just one head, the battle is finished
and  Hercules is the winner
\item Otherwise, Hercules chops off \emph{one} head of the hydra

\begin{itemize}
\item If the head is at distance 1 from the foot, the head is just lost by the hydra
\item Otherwise, let us denote by \(r\) the node that was at distance \(2\) from 
the removed head in the direction of the foot,  and consider the  sub-hydra \(h'\) of \(h\), whose  root is \(r\) \footnote{$h'$ will be called ``the wounded part of the hydra'' in the subsequent text. In Figures~\vref{fig:Hy2} and ~\vref{fig:Hy4}, this sub-hydra  is displayed in red.}. Let $n$ be some natural number.
Then $h'$ is replaced by  $n+1$ of copies of \(h'\) which share the same root $r$.
 The \emph{replication number} $n$ may be different (and generally is)   at each round of the fight.
It may be chosen by the hydra, according to its strategy, or imposed by some 
particular rule. In many presentations of hydra battles, this number is increased by $1$ at each round. In the following presentation, we will also consider battles where the hydra is free to chose its number of replication at every round of the battle\footnote{Let us recall that, if the chopped-off head was at distance 1 from the foot, the replication factor is meaningless.}.
\end{itemize}
\end{itemize}

@@

Moreover, we note that this description is in \emph{imperative} terms. In order to build a formal  study of the properties of hydra battles, we prefer to use mathematical language, i.e., graphs, relations, functions, etc.
Thus, the replication process will be represented as a binary relation on a data type \texttt{Hydra},
linking the state of the hydra \emph{before} and \emph{after} the transformation.
A battle will thus be represented as a sequence of terms of type \texttt{Hydra}.

@@

Next, assume Hercules choses to chop off one of the two highest heads of \texttt{Hy'}, for instance the rightmost one. Fig.~\vref{fig:Hy2} represents the rotten neck in dashed lines, and the part that will be replicated in red. Assume also that the hydra decides to add 4 copies of the red part. We obtain a new state \texttt{Hy''} depicted in Fig.~\ref{fig:Hy3}.

\begin{figure}[h]
\begin{tikzpicture}[very thick, scale=0.6]

@@

\begin{figure}[hp]

\begin{tikzpicture}[very thick, scale=0.5]

\node (foot) at (2,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (2,4)  {{\color{lightred}$\bullet$}};
\node (N3) at (2,6) {{\color{lightred}$\bullet$}};
\node (H0) at (0,2) {$\Smiley[2][vertfluo]$};
\node (H1) at (0,8) {$\Sey[2][lightred]$};
\node (H2) at (5,0) {$\Xey[2][lightgray]$};
\node (H5) at (4,4) {$\Smiley[2][vertfluo]$};
\node (ex) at (5,8) {};
\draw (foot) -- (N1)[very thick] ;
\draw (N1) -- (N2);
\draw  (N2) -- (N3)[draw=lightred];
\draw (N3) to   [bend left= 10](H1) [draw=lightred];
\draw [dashed] (N3) to [bend left= 10](ex);
\draw (foot) to [bend left= 10]  (H0) ;
\draw (N1) to [bend right= 16] (H5);
\end{tikzpicture}
\caption{A beheading}
\label{fig:Hy2}
\end{figure}

\begin{figure}[hp]
\begin{tikzpicture}[very thick, scale=0.6]

@@

\begin{figure}[hp]
\begin{tikzpicture}[very thick, scale=0.6]

@@

\begin{figure}[hp]
\begin{tikzpicture}[very thick, scale=0.4]

@@

Various \emph{Coq} libraries  represent trees the nodes of which 
have a non-fixed number of sons. The reader can, for instance, look at
Chapter 14, pages 400-406 of \cite{BC04}. 
Our representation uses two mutual inductive types: \texttt{Hydra} to represent
 any hydra, and \texttt{Hydrae} to represent finite sequences of hydras.

\index{Types!Hydra}

>>>>

Technically, a \emph{hydra} is just a finite ordered tree, each node of which 
has any number of sons. Note that, contrary to the computer science tradition, we will show the hydras 
with the heads up and the foot (i.e., the root of the tree) down.
Fig.~\ref{fig:Hy} represents such  a hydra, which will be referred to as \texttt{Hy} in our examples (please look at the 
module~\href{../src/html/hydras.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}).

\begin{figure}[h]
\centering
\begin{tikzpicture}[very thick, scale=0.6]

@@

\begin{figure}[h]
  \centering
  \begin{tabular}{ll}
Hydras & Finite rooted trees\\
\hline
foot & root\\
head & leaf\\
node & node\\
segment  & (directed) edge \\
sub-hydra & subtree\\
daughter & immediate subtree\\
\end{tabular}
  \caption{Translation from hydras to trees}
  \label{tab:hyd2tree}
\end{figure}

@@

\label{sec:orgheadline44}
A \emph{hydra battle} is a fight between Hercules and the Hydra. 
More formally, a  battle is a sequence of \emph{rounds}.
At each round:
\begin{itemize}
\item If the hydra is composed of just one head, the battle is finished
and  Hercules is the winner.
\item Otherwise, Hercules chops off \emph{one} head of the hydra,

\begin{itemize}
\item If the head is at distance 1 from the foot, the head is just lost by the hydra, with no more reaction.
\item Otherwise, let us denote by \(r\) the node that was at distance \(2\) from 
the removed head in the direction of the foot,  and consider the  sub-hydra \(h'\) of \(h\), whose  root is \(r\) \footnote{$h'$ will be called ``the wounded part of the hydra'' in the subsequent text. In Figures~\vref{fig:Hy2} and ~\vref{fig:Hy4}, this sub-hydra  is displayed in red.}. Let $n$ be some natural number.
Then $h'$ is replaced by  $n+1$ of copies of \(h'\) which share the same root $r$.
 The \emph{replication number} $n$ may be different (and generally is)   at each round of the fight.
It may be chosen by the hydra, according to its strategy, or imposed by some 
particular rule. In many presentations of hydra battles, this number is increased by $1$ at each round. In the following presentation, we will also consider battles where the hydra is free to chose its number of replication at every round of the battle\footnote{Let us recall that, if the chopped-off head was at distance 1 from the foot, the replication factor is meaningless.}.
\end{itemize}
\end{itemize}

@@

Moreover, we note that this description is in \emph{imperative} terms. In order to build a formal  study of the properties of hydra battles, we prefer to use a mathematical vocabulary, i.e., graphs, relations, functions, etc.
Thus, the replication process will be represented as a binary relation on a data type \texttt{Hydra},
linking the state of the hydra \emph{before} and \emph{after} the transformation.
A battle will thus be represented as a sequence of terms of type \texttt{Hydra}, respecting the rules of the game.

@@

Next, assume Hercules choses to chop off one of the two highest heads of \texttt{Hy'}, for instance the rightmost one. Fig.~\vref{fig:Hy2} represents the rotten neck in dashed lines, and the part that will be replicated in red. Assume also that the hydra decides to add 4 copies of the red part\footnote{In other words, the replication factor at this round is equal to $4$.}. We obtain a new state \texttt{Hy''} depicted in Fig.~\ref{fig:Hy3}.

\begin{figure}[h]
\centering
\begin{tikzpicture}[very thick, scale=0.6]

@@

\begin{figure}[hp]
\centering
\begin{tikzpicture}[very thick, scale=0.5]

\node (foot) at (2,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (2,4)  {{\color{lightred}$\bullet$}};
\node (N3) at (2,6) {{\color{lightred}$\bullet$}};
\node (H0) at (0,2) {$\Smiley[2][vertfluo]$};
\node (H1) at (0,8) {$\Sey[2][lightred]$};
\node (H2) at (5,0) {$\Xey[2][lightgray]$};
\node (H5) at (4,4) {$\Smiley[2][vertfluo]$};
\node (ex) at (5,8) {};
\draw (foot) -- (N1)[very thick] ;
\draw (N1) -- (N2);
\draw  (N2) -- (N3)[draw=lightred];
\draw (N3) to   [bend left= 10](H1) [draw=lightred];
\draw [dashed] (N3) to [bend left= 10](ex);
\draw (foot) to [bend left= 10]  (H0) ;
\draw (N1) to [bend right= 16] (H5);
\end{tikzpicture}
\caption{A second beheading}
\label{fig:Hy2}
\end{figure}

\begin{figure}[hp]
\centering
\begin{tikzpicture}[very thick, scale=0.6]

@@

\begin{figure}[hp]
\centering
\begin{tikzpicture}[very thick, scale=0.6]

@@

\begin{figure}[hp]
\centering
\begin{tikzpicture}[very thick, scale=0.4]

@@

In order to describe trees where each node can have an arbitrary (but finite) number of sons, it usual to define a type where each node carries a \emph{forest}, \emph{i.e} a list of trees
(see for instance Chapter 14, pages 400-406 of \cite{BC04}).

For this purpose, we define two mutual inductive types.
\index{Types!Hydra}

### 2020-07-21 latex/chapter-introduction.tex
corrections in the pdf documentation

This is obviously the case of the four-color theorem~\cite{fourcolors}  and the Kepler conjecture~\cite{flyspeck2015}. We do not mention impressive works like the proof of the odd-order theorem ~\cite{oddorderthm}, since understanding its statement requires some mathematical culture.

Hydra games appeared in an article published in 1982 by two mathematicians:
L. Kirby and J. Paris~\cite{KP82}: \emph{Accessible Independence Results for Peano Arithmetic}. 
Although the mathematical contents of this 
paper are quite advanced, hydra games (a.k.a. \emph{hydra battles}) are very easy to understand. There are now several sites on Internet where you can find tutorials on hydra games, together with simulators you can play with. See, for instance, the page written by Andrej Bauer~\cite{bauer2008}.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=11cm]{epsilon0.jpg}
  \caption{Discrete mathematics and formal proofs}
  \label{fig:gamma0}
\end{figure}
 We also hope to provide highlights on infinity (both potential and actual) through the notions of function, computation, limit,
 types and proofs.

\subsection{Remarks}
In~\cite{KP82}, Kirby and Paris showed  that there is no proof of termination of all hydra battles in Peano Arithmetic (PA).
Since we are used to writing proofs in higher order logic, the restriction to PA was quite unnatural for us. So we chosed to prove another statement without any reference to PA, by considering a class of proofs indexed by ordinal numbers upto $\epsilon_0$.

Unlike mathematical literature, where definitions and proofs are spread over many articles and books,
the whole proof is now inside your computer. It is composed of the \texttt{.v} files you downloaded and 
parts of \coq's standard library. Thus, there is no ambiguity in our definitions and the premises of the theorems. Furthermore, you will be able to navigate through the development, using your favourite editor or IDE, and some commands like \texttt{Search}, \texttt{Locate}, \texttt{Print Assumptions}, etc.

@@

\paragraph{Main references}

In our development, we adapt the definitions and prove many theorems which
we found in the following articles. 
\begin{itemize}
\item ``Accessible independence results for Peano arithmetic''  by Laurie Kirby and Jeff Paris~\cite{KP82}
\item ''Rapidly growing Ramsey Functions'' by Jussi Ketonen and Robert Solovay~\cite{KS81}
\item ``The Termite and the Tower'', by Will Sladek~\cite{Sladek07thetermite}

\subsubsection*{Structure of this document}
\begin{itemize}
\item Whenever several implementations are possible, we will discuss the pros and cons of every possible choice
\item Most of the proofs we present are \emph{constructive}. Whenever possible, we provide the user with an associated function, which she or he can apply in \gallina{} or \ocaml{} in order to get a ``concrete'' feeling of the meaning of the considered theorem.
For instance, in Chapter~\vref{chap:ketonen}, the notion of \emph{limit ordinal} is
made more ``concrete'' thanks to a function \texttt{canonseq} that computes every item of a sequence which converges on a given limit ordinal $\alpha$. This simply typed function allows the user/reader to make her/his own experimentations.
For instance, one can very easily compute the $42$-nd item of a sequence which converges towards $\omega^{\omega^\omega}$.

@@

@@

\subsubsection{Acknowledgements}
\label{sec:orgheadline5}
    Many thanks to David Ilcinkas, Sylvain Salvati, Alan Schmitt and Théo Zimmerman for their help on the elaboration of this document, and to the
 members of the \emph{Formal Methods} team at laBRI for their helpful comments 
on an oral presentation of this work.

@@

\subsubsection{Typographical conventions}

@@

\subsubsection{Alternative or bad definitions}

>>>>

This is obviously the case of the four-color theorem~\cite{fourcolors}  and the Kepler conjecture~\cite{flyspeck2015}. We do not mention impressive works like the proof of the odd-order theorem ~\cite{oddorderthm}, since understanding its statement requires a quite good mathematical culture.

Hydra games (a.k.a. \emph{Hydra battles}) appeared in an article published in 1982 by two mathematicians:
L. Kirby and J. Paris~\cite{KP82}: \emph{Accessible Independence Results for Peano Arithmetic}. 
Although the mathematical contents of this 
paper are quite advanced, the rules of hydra battles are very easy to understand. There are now several sites on Internet where you can find tutorials on hydra games, together with simulators you can play with. See, for instance, the page written by Andrej Bauer~\cite{bauer2008}.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=11cm]{epsilon0.jpg}
  \caption{Discrete mathematics and formal proofs}
  \label{fig:gamma0}
\end{figure}
 We also hope to provide a little clarification on infinity (both potential and actual) through the notions of function, computation, limit,
 types and proofs.

\subsection*{Remarks}
In~\cite{KP82}, Kirby and Paris showed  that there is no proof of termination of all hydra battles in Peano Arithmetic (PA).
Since we are used to writing proofs in higher order logic, the restriction to PA was quite unnatural for us. So we chosed to prove another statement without any reference to PA, by considering a class of proofs indexed by ordinal numbers upto $\epsilon_0$.

Unlike mathematical literature, where definitions and proofs are spread over many articles and books,
the whole proof is now inside your computer. It is composed of the \texttt{.v} files you downloaded and 
parts of \coq's standard library. Thus, there is no ambiguity in our definitions and the premises of the theorems. Furthermore, you will be able to navigate through the development, using your favourite text editor or IDE, and some commands like \texttt{Search}, \texttt{Locate}, \texttt{Print Assumptions}, etc.

@@

\paragraph*{Main references}

In our development, we adapt the definitions and prove many theorems which
we found in the following articles. 
\begin{itemize}
\item ``Accessible independence results for Peano arithmetic''  by Laurie Kirby and Jeff Paris~\cite{KP82}
\item ''Rapidly growing Ramsey Functions'' by Jussi Ketonen and Robert Solovay~\cite{KS81}
\item ``The Termite and the Tower'', by Will Sladek~\cite{Sladek07thetermite}
\item Chapter V of ``Proof Theory'' by Kurt Schütte~\cite{Schutte}
\end{itemize}

\subsubsection*{Structure of this document}
\begin{itemize}
\item Whenever several implementations are possible, we will discuss the pros and cons of every possible choice
\item Most of the proofs we present are \emph{constructive}. Whenever possible, we provide the user with an associated function, which she or he can apply in \gallina{} or \ocaml{} in order to get a ``concrete'' feeling of the meaning of the considered theorem.
For instance, in Chapter~\vref{chap:ketonen}, the notion of \emph{limit ordinal} is
made more ``concrete'' thanks to a function \texttt{canon} which computes every item of a sequence which converges on a given limit ordinal $\alpha$. This simply typed function allows the user/reader to make her/his own experimentations.
For instance, one can very easily compute the $42$-nd item of a sequence which converges towards $\omega^{\omega^\omega}$.

@@

\subsubsection*{State of the development}
The \coq{} scripts herein are in constant development since our contribution~\cite{CantorContrib} on ordinal notations for the ordinals $\epsilon_0$ and $\Gamma_0$.
We added new material : axiomatic definition of countable ordinals after Schütte~\cite{Schutte}, combinatorial aspects of $\epsilon_0$, after Ketonen and Solovay~\cite{KS81} and Kirby and Paris~\cite{KP82}, recent \coq{} technology: type classes, equations, etc.

We are now working in order to make clumsy proofs more readable, simplify definitions, 
to ``factorize'' proofs as much as possible.

Many ``todo''s or projects are put in this text, suggesting possible improvements.

@@

\subsubsection*{Acknowledgements}
\label{sec:orgheadline5}
    Many thanks to David Ilcinkas, Sylvain Salvati, Alan Schmitt and Théo Zimmerman for their help on the elaboration of this document, and to the
 members of the \emph{Formal Methods} team at laBRI for their helpful comments 
on an oral presentation of this work.

@@

\subsection*{Typographical conventions}

@@

\subsection*{Alternative or bad definitions}
